An Efficient Calculus for Generalised Parikh Images of Regular Languages

### 2022-07-06 main.tex
intro

The Parikh map is a characterisation of a context-free language as a vector of
its character counts, famously a semilinear set~\cite{parikh-theorem}. It
appears naturally as part of \Fudge{many operations} in model checking and
solving string constraints in automata-based solvers such as \Ostrich{}
\cite{ostrich}, notably in representing constraints on string lengths. The
enhanced \OstrichPlus{} solver~\cite{ostrich-plus} makes even more extensive use
of Parikh images.

It is possible to compute the Parikh image of any context-free language using
the method described in~\cite{generate-parikh-image}. Later improvements have
produced a construction taking at most linear time to
produce~\cite{muscholl-linear}. However, the resulting existentially quantified
clauses are costly to eliminate as the number of variables increases, in
practice making many real-world problems intractable.

In addition to the elimination of existentially quantified variables, another
problem arises in string solvers in particular. Conjunctions of string
constraints lead to the computation of length images of products (intersections)
of regular languages represented as finite automata. Applying the approach
in~\cite{generate-parikh-image} would in this case require the eager computation
of the product before its length image. In several instances we have observed
while solving real-world string constraints, the computation of the product of
automata exhausts the memory of any machine available due to the exponential
blow-up in size of the product, quickly becoming intractable as the number of
automata in the product increases. The current best published mitigation for
this problem is an over-approximation that works by approximating the Parikh
image of a product of automata to be the conjunction of the image of the
individual automata of the product \cite{approximate-parikh}. This approach only
works for unsatisfiable instances, and comes with a harsh penalty for
satisfiable instances.

>>>>

The Parikh image is a characterisation of formal languages in terms of
their character counts. Given a language over an
alphabet~$\{a_1, \ldots, a_k\}$, the Parikh image is a set of
$k$-dimensional vectors that contains some
vector~$\VectorLiteral{m_1, m_2, \ldots, m_k}$ if and only if the
formal language contains a word in which each $a_i$ occurs $m_i$
times. It is a classical result that the Parikh image of every
context-free language (and, thus, also of every regular language) is a
semilinear set~\cite{parikh-theorem}, i.e., Presburger-definable.

Parikh images play a central role in many automata-based algorithms,
for instance and notably in today's string solvers, which often have
to process constraints that combine regular language membership with
word length. To decide whether a simple formula like
$x \in \mathcal{L}_1 \wedge y \in \mathcal{L}_2 \wedge |x| > |y|$,
with string variables~$x, y$ and regular
languages~$\mathcal{L}_1, \mathcal{L}_2$, is satisfiable, it is
necessary to reason about the sets of word lengths induced by
$\mathcal{L}_1, \mathcal{L}_2$, which is a special case of the Parikh
image.  This required combined reasoning about strings and string
length has long been identified as a major bottleneck in string
solvers
\cite{DBLP:conf/cav/AbdullaACHRRS15,length-aware-solver,approximate-parikh,more}.
Other string solvers make use of Parikh automata~\cite{X}, and thus
Parikh images in the general case, to handle operations including
\verb!str.substr!  and \verb!str.at!, which comes at an even higher
price in terms of computational complexity~\cite{ostrich-plus}.

\iffalse
It appears naturally as part of \Fudge{many operations} in model
checking and solving string constraints in automata-based solvers such
as \Ostrich{} \cite{ostrich}, notably in representing constraints on
string lengths. The enhanced \OstrichPlus{} solver~\cite{ostrich-plus}
makes even more extensive use of Parikh images.
\fi

It is possible to compute an existential Presburger formula describing
the Parikh image of any context-free language in linear
time~\cite{generate-parikh-image}; for the special case of regular
languages, this result was also stated in
\cite{muscholl-linear}. While theoretically elegant, this construction
has several disadvantages, often making it unpractical for integration
into algorithms. Firstly, the constructed Presburger formula contains
a linear number of existential quantifiers in the size of the
considered grammar, as well as complex Boolean structure, which is
needed to express the connectedness of sets of productions considered
in the construction. Eliminating those quantifiers to obtain a
quantifier-free representation of the Parikh image has exponential
complexity~\cite{X}, and is in practice often impossible in reasonable
time. Just solving the Parikh image membership problem is NP-complete,
as it corresponds to computing a satisfying assignment of the
existential Presburger formula, and taxing for solvers as
well~\cite{ostrich-plus}.

\iffalse
Later improvements have produced a construction taking at
most linear time to produce~\cite{muscholl-linear}. However, the
resulting existentially quantified clauses are costly to eliminate as
the number of variables increases, in practice making many real-world
problems intractable.
\fi

Secondly, in applications involving regular languages, it is typically
necessary to consider the Parikh image not only of a single automaton,
but of the intersection of multiple automata. This problem arises in
string solvers in particular, as conjunctions of string constraints
lead to the computation of length images of products (intersections)
of regular languages represented as finite automata. Applying the
approach in~\cite{generate-parikh-image} would in this case require
the eager computation of the product before its length image, and
result in an existential Presburger formula of exponential size (in
the number of automata). In several instances we have observed while
solving real-world string constraints, the computation of the product
of automata exhausts the memory of any machine available due to the
exponential blow-up in size of the product, quickly becoming
intractable as the number of automata in the product increases. The
current best published mitigation for this problem is an
over-approximation that works by approximating the Parikh image of a
product of automata to be the conjunction of the image of the
individual automata of the product \cite{approximate-parikh}. This
approach only works for unsatisfiable instances, and comes with a
harsh penalty for satisfiable instances.
